Finiteness in Cubical Type Theory

### 2020-01-25 paper.tex
more

The finiteness predicates we are interested in are organised in
Figure~\ref{finite-classification}.
We will explore two aspects of each predicate: its relation to the other
predicates, and its topos-theoretic classification.

When we say ``relation'' we are referring to the arrows in
Figure~\ref{finite-classification}.
The proofs, then, amount to a function which inhabits each arrow.
Each unlabelled arrow is a weakening: i.e., every manifest Bishop finite set
is manifest enumerable.
The labelled arrows are strengthening proofs: i.e., every manifest enumerable set
\emph{with decidable equality} is manifest Bishop finite.
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.

We will then characterise each predicate as some form of topos.
Our proofs follow the structure of \cite[Chapters 9, 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.
This means we will define first the precategory of sets, and then the category
of sets, and prove the required closures and limits to get us to a topos.
We say ``some form of'' topos here because of course the category of sets in
HoTT do \emph{not}, in fact, form a topos, rather a \(\Pi W\)-pretopos.
Our main result here is that the category of decidable Kuratowski finite sets
forms a \(\Pi\)-pretopos.

After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates.
We here show that countably finite sets form a \(W\)-pretopos.

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in the
final section, where we implement a library for proof search, based on
omniscience and exhaustibility.
While proof search based on finiteness is not new, implementing it here does
give us an opportunity to demonstrate some actual \emph{computation} in a
univalent setting.
Many of the finiteness predicates are built using univalence, so for the proof
search functionality to work at all we require computational content in the
univalence axiom.
Furthermore, the extensionality afforded to us by HoTT means that we can have
\(\Pi\) types in the domain of the search, meaning we can synthesise full
functions from specifications alone.
\subsection{Related Work}
\subsection{Notation and Background}
\subsubsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into type.
  \item[Universes]
    Implicitly, types will exist at some point in a universe hierarchy,
    beginning with \(\mathbf{Type}\) (\(\mathbf{Type} : \mathbf{Type}_1\),
    \(\mathbf{Type}_1 : \mathbf{Type}_2\), and so on).
    In our formalisation, every proof is done in the most universe polymorphic
    way possible: the proof that the cardinally finite sets form a
    \(\Pi\)-pretopos, for instance, is defined over any universe level.
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] Disjoint union can be defined in terms of \(\Sigma\):
    \begin{equation}
      A \uplus B \coloneqq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then } A \text{ else } B
    \end{equation}
    However, we prefer to use it as an inductively defined type (though the two
    are equivalent).
    \begin{equation}
      \begin{aligned}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B
      \end{aligned}
    \end{equation}
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage.
\end{description}
\subsubsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, rather than an axiom.
This allows us to actually compute with univalent proofs, a capability missing
from HoTT.
\begin{romdefinition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{romdefinition}

@@

  It is defined as a Higher Inductive Type:
  \begin{equation} {
    \begin{alignat*}{3}
      \lVert A \rVert \coloneqq & \; \lvert \wc \rvert &&: A \rightarrow \lVert A \rVert ; \\
                              | & \; \text{squash}     &&: \Pi {(x, y : \lVert A \rVert)} , x \equiv y  ; 
    \end{alignat*} }
  \end{equation}
  We will use three eliminators from \(\lVert A \rVert\) in this paper.
  \begin{enumerate}
  \item \label{elim-prop-prop} For any function \(A \rightarrow B\), where
    \(\text{isProp}(B)\), we have a function \(\lVert A \rVert \rightarrow B\).
  \item \label{elim-prop-monad} A special case of \ref{elim-prop-prop} implies
    that \(\lVert \cdot \rVert\) forms a monad: this means that we get the usual
    Monadic operators on \(\lVert \cdot \rVert\) (bind, pure, fmap, etc.).
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\) (\(\Pi {(x , y : A)} , f(x) \equiv f(y) \)).
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{romdefinition}
\begin{romdefinition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignat}{5}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
                  | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{romdefinition}
\section{Finiteness Predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
The reason we explore predicates other than our focus (cardinal finiteness) is
that we can often prove things like closure much more readily on the simpler
predicates.
The relations (which we will prove in the next section) then allow us to
transfer those proofs onto Kuratowski finiteness.
\subsection{Split Enumerability}
\begin{romdefinition}[Split Enumerable Set] \label{split-enum-def}
  \begin{equation} \label{split-enum-def-eqn}
    \mathcal{E}!(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{romdefinition}

@@

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{romlemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{romlemma}
\begin{romlemma}
  This circle \(S^1\) is Kuratowski finite.
\end{romlemma}
The second of these in particular tells us that the ``Kuratowski-finite types''
are not necessarily sets; it also tells us that we cannot derive decidable
equality from a proof of Kuratowski finiteness.
\section{Relations Between Each Finiteness Definition}
We now look at under what conditions a given finiteness predicate satisfies
another.
\subsection{Split Enumerability and Manifest Bishop Finiteness}
While manifest Bishop finiteness might seem stronger than split enumerability,
it turns out this is not the case.
Both types imply the other.
\begin{romlemma}
  Any manifest Bishop finite type is split enumerable.
\end{romlemma}
\begin{proof}
  The only difference between \(\mathcal{B}\) and \(\mathcal{E}!\) is the
  membership term.
  To derive \(\mathcal{E}!\) from \(\mathcal{B}\), then, we need to derive
  \(\in\) from \(\inunique\).
  We can do this with fst.
  \begin{alignat*}{4}
    &\text{fst} : && \; \Sigma(x : A) , U(x) \rightarrow A \\
    && \simeq      & \; \text{isContr}(A) \rightarrow A \\
    && \simeq      & \; (x \in \mathit{xs}) \rightarrow (x \inunique \mathit{xs})
  \end{alignat*}
  \qed
\end{proof}

To derive \(\mathcal{B}\) from \(\mathcal{E}!\) takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, all we need to show is that the membership proof carries over
appropriately.
\begin{romlemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Manifest Enumerability}
\begin{romlemma}
  Any manifestly Bishop finite type is manifestly enumerable.
\end{romlemma}
\begin{romlemma}
  A manifestly enumerable type with decidable equality is manifest Bishop
  finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{romlemma}
  Any manifest Bishop finite type is cardinal finite.
\end{romlemma}
\begin{romtheorem}
  Any cardinal finite type with a total order is Bishop finite.
\end{romtheorem}
\section{Topos}
Our main result in this section is the formal proof that decidable Kuratowski
finite types form a \(Pi\)-pretopos.
As we saw in theorem~\ref{kuratowski-cardinal}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first two closure proofs, we only consider manifest Bishop finiteness:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.
\begin{romlemma}
  \(\bot\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  From lemma~\ref{bishop-equiv} we can prove the goal by constructing a term of
  the type:
  \begin{equation}
    \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq \bot)
  \end{equation}
  To do so, we take \(n\) to be 0, and by definition~\ref{Fin},
  \(\mathbf{Fin}(0) = \bot\).
  The equivalence is simply the identity equivalence.
\end{proof}
\begin{romlemma}
  \(\top\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  Again, we need only provide a suitable cardinality (in this case 1), and an
  equivalence with \(\mathbf{Fin}(1)\) to prove our goal.
  \begin{align*}
    \top &\simeq \mathbf{Fin}(1) \\
          &\simeq \top \uplus \mathbf{Fin}(0) \\
          &\simeq \top \uplus \bot \\
          &\simeq \top
  \end{align*}
  \qed
\end{proof}
\begin{romlemma}
  \(\mathbf{Bool}\) is manifest Bishop finite.
\end{romlemma}

@@

We have proven the above statements for both \(\mathit{Set}\) and
\(\mathit{FinSet}\).
As far as we know, this is the first formalisation of either.
\section{Countably Infinite Types}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Two Countable Types}
The two types for countability we will consider are analogous to split
enumerability and cardinal finiteness.
The change will be a simple one: in terms of surjections and bijections, we will
swap out \(\mathbf{Fin}\) for \(\mathbb{N}\).
In terms of support lists and cover proofs, we will swap out lists for streams.
A stream can be defined many ways in type theory: we take rather a low-tech
approach here, saying that a stream is a function from the natural numbers.
\begin{romdefinition}[Stream]
  \begin{equation}
    \mathbf{Stream}(A) \coloneqq \mathbb{N} \rightarrow A
  \end{equation}
  Streams are also definable as containers:
  \begin{equation}
    \mathbf{Stream}(A) \simeq \llbracket \top , \text{const}(\mathbb{N}) \rrbracket
  \end{equation}
  However the two definitions are clearly equivalent, and furthermore it is
  simpler to work with the first.
\end{romdefinition}

@@

From both of these types we can derive decidable equality.
\begin{romlemma}
  Any countable type has decidable equality.
\end{romlemma}
\subsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
\subsection{Categories}
\section{Search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

>>>>

The finiteness predicates we are interested in are organised in
figure~\ref{finite-classification}, and defined in
section~\ref{finiteness-predicates}.
We will explore two aspects of each predicate: its relation to the other
predicates, and its topos-theoretic classification.

When we say ``relation'' we are referring to the arrows in
figure~\ref{finite-classification}.
In section~\ref{relations} we will provide a function which inhabits each arrow.
Each unlabelled arrow is a weakening: i.e., every manifest Bishop finite set
is manifest enumerable (lemma~\ref{manifest-bishop-to-manifest-enum}).
The labelled arrows are strengthening proofs: i.e., every manifest enumerable set
\emph{with decidable equality} is manifest Bishop finite
(lemma~\ref{manifest-enum-to-manifest-bishop}).
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.

We will then examine the closure properties of each predicate in
section~\ref{topos}, culminating in a proof that decidable Kuratowski finite
sets form a \(\Pi\)-pretopos (theorem~\ref{finite-topos}).
Our proofs follow the structure of \cite[Chapters 9, 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.

After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates
(section~\ref{infinite}).

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in
section~\ref{search}, where we implement a library for proof search, based on
omniscience and exhaustibility.
While proof search based on finiteness is not new, implementing it here does
give us an opportunity to demonstrate some actual \emph{computation} in a
univalent setting.
Many of the finiteness predicates are built using univalence, so for the proof
search functionality to work at all we require computational content in the
univalence axiom.
Furthermore, the extensionality afforded to us by HoTT means that we can have
\(\Pi\) types in the domain of the search, making our library more powerful than
comparable libraries.
We will demonstrate some interesting uses for this feature in particular.
\subsection{Notation and Background}
\subsubsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into type.
  \item[Universes]
    Implicitly, types will exist at some point in a universe hierarchy,
    beginning with \(\mathbf{Type}\) (\(\mathbf{Type} : \mathbf{Type}_1\),
    \(\mathbf{Type}_1 : \mathbf{Type}_2\), and so on).
    In our formalisation, every proof is done in the most universe polymorphic
    way possible: the proof that the cardinally finite sets form a
    \(\Pi\)-pretopos, for instance, is defined over any universe level.
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] Disjoint union can be defined in terms of \(\Sigma\):
    \begin{equation}
      A \uplus B \coloneqq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then } A \text{ else } B
    \end{equation}
    However, we prefer to use it as an inductively defined type (though the two
    are equivalent).
    \begin{equation}
      \begin{aligned}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B
      \end{aligned}
    \end{equation}
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage.
\end{description}
\subsubsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, rather than an axiom.
This allows us to actually compute with univalent proofs, a capability missing
from HoTT.
\begin{romdefinition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{romdefinition}

@@

  It is defined as a Higher Inductive Type:
  \begin{equation} {
    \begin{alignat*}{3}
      \lVert A \rVert \coloneqq & \; \lvert \wc \rvert &&: A \rightarrow \lVert A \rVert ; \\
                              | & \; \text{squash}     &&: \Pi {(x, y : \lVert A \rVert)} , x \equiv y  ; 
    \end{alignat*} }
  \end{equation}
  We will use three eliminators from \(\lVert A \rVert\) in this paper.
  \begin{enumerate}
  \item \label{elim-prop-prop} For any function \(A \rightarrow B\), where
    \(\text{isProp}(B)\), we have a function \(\lVert A \rVert \rightarrow B\).
  \item \label{elim-prop-monad} A special case of \ref{elim-prop-prop} implies
    that \(\lVert \cdot \rVert\) forms a monad: this means that we get the usual
    Monadic operators on \(\lVert \cdot \rVert\) (bind, pure, fmap, etc.).
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\) (\(\Pi {(x , y : A)} , f(x) \equiv f(y) \)).
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{romdefinition}
\begin{romdefinition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignat}{5}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
                  | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{romdefinition}
\section{Finiteness Predicates} \label{finiteness-predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
The reason we explore predicates other than our focus (cardinal finiteness) is
that we can often prove things like closure much more readily on the simpler
predicates.
The relations (which we will prove in the next section) then allow us to
transfer those proofs onto Kuratowski finiteness.
\subsection{Split Enumerability}
\begin{romdefinition}[Split Enumerable Set] \label{split-enum-def}
  \begin{equation} \label{split-enum-def-eqn}
    \mathcal{E}!(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{romdefinition}

@@

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{romlemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{romlemma}
\begin{romlemma}
  This circle \(S^1\) is Kuratowski finite.
\end{romlemma}
The second of these in particular tells us that the ``Kuratowski-finite types''
are not necessarily sets; it also tells us that we cannot derive decidable
equality from a proof of Kuratowski finiteness.
\section{Relations Between Each Finiteness Definition} \label{relations}
We now look at under what conditions a given finiteness predicate satisfies
another.
\subsection{Split Enumerability and Manifest Bishop Finiteness}
While manifest Bishop finiteness might seem stronger than split enumerability,
it turns out this is not the case.
Both types imply the other.
\begin{romlemma} \label{manifest-bishop-to-split-enum}
  Any manifest Bishop finite type is split enumerable.
\end{romlemma}

To derive \(\mathcal{B}\) from \(\mathcal{E}!\) takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, all we need to show is that the membership proof carries over
appropriately.
\begin{romlemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Manifest Enumerability}
\begin{romlemma} \label{manifest-bishop-to-manifest-enum}
  Any manifestly Bishop finite type is manifestly enumerable.
\end{romlemma}
\begin{romlemma} \label{manifest-enum-to-manifest-bishop}
  A manifestly enumerable type with decidable equality is manifest Bishop
  finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{romlemma} \label{manifest-bishop-to-cardinal}
  Any manifest Bishop finite type is cardinal finite.
\end{romlemma}
\begin{romtheorem} \label{cardinal-to-manifest-bishop}
  Any cardinal finite type with a total order is Bishop finite.
\end{romtheorem}
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\begin{romlemma} \label{cardinal-kuratowski}
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}(A) \times \text{Discrete}(A)
  \end{equation}
\end{romlemma}
\section{Topos} \label{topos}
Our main result in this section is the formal proof that decidable Kuratowski
finite types form a \(Pi\)-pretopos.
As we saw in theorem~\ref{kuratowski-cardinal}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first two closure proofs, we only consider manifest Bishop finiteness:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.
\begin{romlemma}
  \(\bot\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  From lemma~\ref{bishop-equiv} we can prove the goal by constructing a term of
  the type:
  \begin{equation}
    \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq \bot)
  \end{equation}
  To do so, we take \(n\) to be 0, and by definition~\ref{Fin},
  \(\mathbf{Fin}(0) = \bot\).
  The equivalence is simply the identity equivalence.
\end{proof}
\begin{romlemma}
  \(\top\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  Again, we need only provide a suitable cardinality (in this case 1), and an
  equivalence with \(\mathbf{Fin}(1)\) to prove our goal.
  \begin{align*}
    \top &\simeq \mathbf{Fin}(1) \\
          &\simeq \top \uplus \mathbf{Fin}(0) \\
          &\simeq \top \uplus \bot \\
          &\simeq \top
  \end{align*}
  \qed
\end{proof}
\begin{romlemma}
  \(\mathbf{Bool}\) is manifest Bishop finite.
\end{romlemma}

@@

We have proven the above statements for both \(\mathit{Set}\) and
\(\mathit{FinSet}\).
As far as we know, this is the first formalisation of either.
\begin{romtheorem} \label{finite-topos}
  The category of finite sets, \(\mathit{FinSet}\), forms a \(\Pi\)-topos.
\end{romtheorem}
\section{Countably Infinite Types} \label{infinite}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Two Countable Types}
The two types for countability we will consider are analogous to split
enumerability and cardinal finiteness.
The change will be a simple one: in terms of surjections and bijections, we will
swap out \(\mathbf{Fin}\) for \(\mathbb{N}\).
In terms of support lists and cover proofs, we will swap out lists for streams.
A stream can be defined many ways in type theory: we take rather a low-tech
approach here, saying that a stream is a function from the natural numbers.
\begin{romdefinition}[Stream]
  \begin{equation}
    \mathbf{Stream}(A) \coloneqq \mathbb{N} \rightarrow A
  \end{equation}
  Streams are also definable as containers:
  \begin{equation}
    \mathbf{Stream}(A) \simeq \llbracket \top , \text{const}(\mathbb{N}) \rrbracket
  \end{equation}
  However the two definitions are clearly equivalent, and furthermore it is
  simpler to work with the first.
\end{romdefinition}

@@

From both of these types we can derive decidable equality.
\begin{romlemma}
  Any countable type has decidable equality.
\end{romlemma}
\subsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
\subsection{Categories}
\section{Search} \label{search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.
